D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-21 ootavacuous.tex
Tie Section 8.2 discussion to quasi-volatile properties
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

The following sections discuss these constraints, starting with the
laws of physics, continuing with hardware constraints, continuing
further with constraints imposed by the standard, and culminating
with constraints imposed by the C++ implementations themselves.
An additional section discusses the OOTA-cycle implications for
any tooling that exists separately for these C++ implementations.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
memory model.
In particular, implementations must not behave as if the source code has
additional:

\begin{enumerate}
\item	Invented atomic stores.
\item	Duplicated atomic stores.
\end{enumerate}

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted and on fusing of
nonvolatile atomic loads from adjacent objects.

@@

Duplicating atomic stores can also introduce new and undesirable
behaviors.
To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
litmus test that demonstrates such a behavior.
Without line~9, if the value loaded into \co{r1} is 1, then the final value
of \co{x} must be 2.
With line~9, the final value of \co{x} can be 1 even when the
value loaded into \co{r1} is 1.
The implementation is therefore forbidden from duplicating that store of 1
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
Again, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

>>>>

The following sections discuss these constraints, starting with
with hardware constraints, continuing
further with constraints imposed by the standard, and culminating
with constraints imposed by the C++ implementations themselves.
An additional section discusses the OOTA-cycle implications for
any tooling that exists separately for these C++ implementations.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
memory model.
The ``order-preserving'' bullet from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering only of quasi volatile accesses to the same object,
but such reordering is in any case directly forbidden by the standard.
Referring also to the ``onto'' bullet from
Section~\ref{sec:Volatile and Quasi Volatile Accesses},
implementations must not behave as if the source code has
additional:

\begin{enumerate}
\item	Invented atomic stores.
\item	Duplicated atomic stores.
\item	Invented atomic loads.
\end{enumerate}

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted and on fusing of
nonvolatile atomic loads from adjacent objects.
These topics correspond to the ``one-to-one'' and ``total'' bullets of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

@@

Duplicating atomic stores is a special case of inventing them, but it is
worth illustrating the fact that duplicating stores can also introduce
new and undesirable behaviors.
To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
litmus test that demonstrates such a behavior.
Without line~9, if the value loaded into \co{r1} is 1, then the final value
of \co{x} must be 2.
With line~9, the final value of \co{x} can be 1 even when the
value loaded into \co{r1} is 1.
The implementation is therefore forbidden from duplicating that store of 1
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
Again, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

\paragraph{Invented Atomic Loads}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
excessive cache-miss overheads.

\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Do Not Duplicate Atomic Loads}
\label{lst:Do Not Duplicate Atomic Loads}
\end{listing}

However, even duplicating an existing atomic load can lead to incorrect
results.
To see this, please refer to
Listing{lst:Do Not Duplicate Atomic Loads}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} might
not be prepared to deal with.

Therefore, C++ implementations should not invent loads, and should
not even duplicate them.
